---
title: "Cyclistic bike share - Capstone Project"
output: github_document
---
# Case Study 
## About the company
### Cyclistic, a bike-share company in Chicago (fictional).
##The director of marketing.
#### Lily Moreno: 

Cyclistic marketing analytics team:A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy.I joined this team six months ago.

Cyclistic executive team: Detail-oriented executive team will decide whether to approve the recommended marketing program.

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.

Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships.

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, The director of marketing believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers,

Moreno believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.

Moreno has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends.

### Steps of the data analysis process:
#### ask
#### prepare
#### process
#### analyze
#### share
#### act

# 1.ASK

## Guiding questions:

#### What is the problem you are trying to solve? Maximize the number of annual memberships.
#### Why would casual riders buy Cyclistic annual memberships?
#### How can Cyclistic use digital media to inﬂuence casual riders to become members?
#### How can your insights drive business decisions?
#### How do annual members and casual riders use Cyclistic bikes diferently?
#### What makes casual riders become annual members?


##Business Task
##How to Convert casual riders into Cyclistic members?

## key stakeholders:
### Cyclistic Executive team
### Cyclistic Marketing Analytics team
### Cyclistic users (Casual Riders, Cyclistic members)


# PREPARE
To analyze and identify trends,i wil use the previous 12 months of Cyclistic’s historical trip data, downloaded here <https://divvy-tripdata.s3.amazonaws.com/index.html>

Previous 12 months .cvs files trip data, from 2022/01 to 2022/12
13 variables
Note: The datasets have a different name because Cyclistic is a fictional company.

#This data is ROCCC:
Reliable - The data has been made available by Motivate International Inc. under this license.
Original- Lyft Bikes and Scooters, LLC, operates the City of Chicago’s, subject to the terms and conditions of the Data License Agreement.
Comprehensive- The data is organized in .cvs files by trip month and by trip quarter.
Corrent- This data is updated monthly. I will use the previous 12 months, from January 2022 to December 2022.
Cited- The dataset is public and available here <https://divvy-tripdata.s3.amazonaws.com/index.html>









```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Install and load the various packages used in this analysis
```{r}

installed.packages("tidyverse")
installed.packages("janitor")
installed.packages("dplyr")    # alternative installation of the %>%
library(ggplot2)
library(magrittr) # needs to be run every time you start R and want to use %>%
# alternatively, this also loads 
library(tidyverse)
library(dplyr) 
library(janitor)
```
```{r}
rm(list = ls())
```

# Reading the previous 12 months of Cyclistic trip data

```{r}
df1 <- read.csv("./data/202201.csv")
df2 <- read.csv("./data/202202.csv")
df3 <- read.csv("./data/202203.csv")
df4 <- read.csv("./data/202204.csv")
df5 <- read.csv("./data/202205.csv")
df6 <- read.csv("./data/202206.csv")
df7 <- read.csv("./data/202207.csv")
df8 <- read.csv("./data/202208.csv")
df9 <- read.csv("./data/202209.csv")
df10 <- read.csv("./data/202210.csv")
df11 <- read.csv("./data/202211.csv")
df12 <- read.csv("./data/202212.csv")
```

#combine 12 data frames into one date frame

```{r}
bike_rides <- rbind(df1,df2, df3, df4, df5, df6, df7, df8, df9, df10, df11, df12)
bike_rides <- janitor :: remove_empty(bike_rides, which = c("cols"))
bike_rides <- janitor :: remove_empty(bike_rides, which = c("rows"))
library(dplyr)
bike_rides <- bike_rides %>% dplyr::filter(bike_rides$start_station_name !="")
bike_rides <- bike_rides %>% dplyr::filter(bike_rides$end_station_name !="")
bike_rides <- bike_rides %>% dplyr::filter(bike_rides$end_station_id !="")

```


#PROCESS
Transform data in order to work effectively

# Convert Data/Time stamp to Data/Time....
##change  data type format of columns "started_at" and "ended_at" to datetime in order to get consistency

```{r}
bike_rides$Ymd <- as.Date(bike_rides$started_at)
bike_rides$started_at <- lubridate::ymd_hms(bike_rides$started_at) 
bike_rides$ended_at <- lubridate::ymd_hms(bike_rides$ended_at)

bike_rides$start_hour <- lubridate::hour(bike_rides$started_at)
bike_rides$end_hour <- lubridate::hour(bike_rides$ended_at)
```

# Difference between start and end hours

```{r}


bike_rides$Hours <- difftime(bike_rides$ended_at, bike_rides$started_at, unit = c("hours"))
bike_rides$Minutes <- difftime(bike_rides$ended_at, bike_rides$started_at, unit = c("mins"))

bike_rides <- bike_rides %>% dplyr::filter(Minutes >0)
```

```{r}
installed.packages("skimr")
library(skimr)
skim_without_charts(bike_rides)
```


#Create summary of data frame


```{r}

library(dplyr)
bikesrides2 <- bike_rides %>% group_by(Weekly = 
                                       lubridate::floor_date(Ymd, "week"),start_hour) %>%
  summarise(
    Minutes = sum(Minutes),
    Mean = mean(Minutes),
    Median = median(Minutes),
    Maximun = max(Minutes),
    Minimum = min(Minutes),
    Count = n(),
   ## .groups = 'drop'
  ) %>% ungroup()
```

```{r}
##install.packages('forecast', dependencies = TRUE)
```


```{r}
bikesrides2$CntMA <- forecast::ma(bikesrides2$Count,28)
```


# Plot of Rides By Date
### Summary Stats : Counts



# Summary of Hourly Counts



```{r}
summary(bikesrides2$Count)
```

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
17.0   834.5  2454.5  3800.4  5826.8 16930.0 

# Table of Counts by Hour


```{r}
xtabs(bikesrides2$Count~bikesrides2$start_hour)
```
bikesrides2$start_hour
0       1        2        3        4 
67610   42585   25110    14895    12896 
5       6        7        8        9 
37994  105506   194801   238902   185283 
10      11       12       13       14 
195349  242821  282872   284686   291629 
15      16       17       18       19 
339664  422100  496697   416059   303790 
20      21       22       23 
215154  174165  143531   100049 


#ANALYSE


```{r}
bikesrides2$Monthly <- lubridate::month(bikesrides2$Weekly)

bikesrides2 %>% ggplot() + geom_col(mapping=aes(x=Weekly, y=Count)) +
  scale_y_continuous(breaks=c(1),labels = ",") +
  labs(title = "Count of Rides per Day" ,
  subtitle  = "Bases on 28 day moving average",
   y="Average rides per day")
```

```{r}
bikesrides2 %>%  
  ggplot()+geom_col(aes(x=start_hour, y=Count))+
  scale_y_continuous(breaks = c(1), labels = ",") +
  labs(title = "Count of Rides per hours",
       y="Rides pe Hour")
```
#Number of rides and average ride length segmented by rider type

```{r}
bike_rides %>% 
  group_by(member_casual) %>% 
  summarise(Average_ride_length = mean(Hours)) %>% 
  ggplot(mapping = aes(x = member_casual, y = Average_ride_length, fill = member_casual)) + geom_col() + labs(title = "Average ride length by Rider type", x="Rider type", y="Average ride length")
```


#Count of Rides by bike Type
## Summary of bike Types

 
```{r}
 bikestype <- bike_rides %>% 
   group_by(rideable_type, Weekly=
              lubridate::floor_date(Ymd,"week")) %>% 
   summarise(
     Minutes = sum(Minutes),
     Mean = mean(Minutes),
     Median = median(Minutes),
     Maximun = max(Minutes),
     Minimum = min(Minutes),
     Count = n()
   ) %>% ungroup()
```



#Count by Bike Type(total by Week)
 
```{r}
 table(bikestype$rideable_type)
```
  
  
  classic_bike   docked_bike electric_bike 
  53            53            53
```{r}  
  library(ggplot2)
```
  
  # Number of rides weekly and ride type
```{r}
  ggplot(bikestype)+geom_area(aes(x=Weekly, y=Count,fill=rideable_type))
```

  
  # TOP 20 start stations name and number if rides on these stations
  
```{r}
bike_rides %>%  count(start_station_name,sort = TRUE) %>% 
  top_n(20) %>% ggplot()+geom_col(aes(x=reorder(start_station_name,n), y = n))+
  coord_flip()+labs(title = "top 20 start stations by Ride Count",
                    y="Count of Rides",x="Station Name")

scale_y_continuous(labels=" ")
```






